Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-23 paper.tex

The output of population genetics simulators, quite naturally,
tends to be some representation of the resulting nucleotide sequences
(or some derived statistics of those sequences),
although the underlying genealogies are also of interest.
For instance,
the classic simulator \ms\ will output trees or genome sequences,
but not both.
% Mutations are not treated as first-class data --- the sequences
% were really all people were after, classically.
However, joint information about the mutations that have been simulated
-- e.g., where they have occurred on the trees -- 
is usually not retained or made available for subsequent analysis.

@@

% Explain the background: what have people done classically?
Because coalescent simulations are usually concerned with
neutral evolution (see the \nameref{sec-selection} section, however)
the problem of generating synthetic genetic variation can be decomposed into
two independent steps:
simulating genetic ancestry (the trees), and then simulating
variation by superimposing mutation processes on those trees
(see figure~\ref{fig-mutated-trees}).
Indeed, to call a mutation ``neutral'' means that its presence or absence
does not affect demography.
A number of programs exist to place mutations on trees:
in addition to \phastsim~\citep{demaio2021phastsim},
the classical \texttt{Seq-Gen} program~\citep{rambaut1997seq}
allows a range of different models of sequence evolution to be imposed on a given
tree in Newick format, and there are various
extensions to the basic nucleotide
models~\citep[e.g.][]{cartwright2005dna,fletcher2009indelible}.
Population genetic simulators have tended to
include their own implementations of mutation simulation, with
most just supporting the infinite sites
model~\citep[e.g.][]{hudson2002generating}
but with several supporting a range of different models
of sequence
evolution~\citep[e.g.][]{mailund2005coasim,excoffier2011fastsimcoal,
virgoulay2021gspace}.

>>>>

The output of population genetics simulators, quite naturally,
tends to be some representation of the resulting nucleotide sequences
(or some derived statistics of those sequences),
although the underlying genealogies are also of interest.
% Mutations are not treated as first-class data --- the sequences
% were really all people were after, classically.
However, joint information about the mutations that have been simulated
-- e.g., where they have occurred on the trees -- 
is usually not retained or made available for subsequent analysis.

@@

% Explain the background: what have people done classically?
Because coalescent simulations are usually concerned with
neutral evolution (see the \nameref{sec-selection} section, however)
the problem of generating synthetic genetic variation can be decomposed into
two independent steps:
simulating genetic ancestry (the trees), and then simulating
variation by superimposing mutation processes on those trees
(see figure~\ref{fig-mutated-trees}).
A number of programs exist to place mutations on trees:
for instance,
the classical \texttt{Seq-Gen} program~\citep{rambaut1997seq}
allows a range of different models of sequence evolution to be imposed on a given
tree in Newick format, and there are various
extensions to the basic nucleotide
models~\citep[e.g.][]{cartwright2005dna,fletcher2009indelible}.
Population genetic simulators have tended to
include their own implementations of mutation simulation, with
most just supporting the infinite sites
model~\citep[e.g.][]{hudson2002generating}
but with several supporting a range of different models
of sequence
evolution~\citep[e.g.][]{mailund2005coasim,excoffier2011fastsimcoal,
virgoulay2021gspace}.
